Windowed least-squares model reduction for dynamical systems

### 2019-10-06 main.tex
fix eq 10
math thru 2.3.1

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG projection. Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
framework. Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods. Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs.
Section~\ref{sec:numerical_experiments} presents numerical experiments.
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with italicized bold symbols (e.g., $\boldsymbol
x$), vectors with standard bold symbols (e.g., $\mathbf{x}$), 
matrices with capital bold symbols (e.g., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic symbols (e.g., $\mathcal{X}$). We additionally denote differentiation of a time-dependent 
function with respect to time as $\stateDot \equiv \frac{d \state}{dt}$.
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical formulation}\label{sec:math}
	We begin by providing the formulation for the full-order model,
	followed by a description of standard model-reduction methods
	classified according to the type of trial subspace they employ.
	\subsection{Full-order model}\label{sec;FOM}
We consider the full-order model to be a dynamical system expressed as a
	system of ordinary differential equations (ODEs)
\begin{equation}\label{eq:FOM}
 \stateFOMDotArg{t}  = \velocity(\stateFOMArg{}{t},t), \qquad \stateFOM(0) =
	\stateFOMIC,\qquad t \in [0,T]
\end{equation}
where $\stateFOMDot \equiv \frac{d \stateFOM}{dt}$, 
$T \in \mathbb{R}^+$ denotes the final time, $\stateFOM \in \RR{\fomdim}
	\otimes \timeSpace $ with $\stateFOM: [0,T] \rightarrow  \RR{\fomdim}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, 
	$\timeSpace$ denotes the set of real-valued functions acting on the time
	domain, 
 $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial condition,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,\timeDumArg) \mapsto
	\velocity(\stateyDiscrete,\timeDumArg)$ denotes the velocity, which is possibly
	nonlinear in its first argument. We refer to Eq.~\eqref{eq:FOM} 
	as the ``full-order model" (FOM) ODE.

Directly solving initial value problem~\eqref{eq:FOM} is computationally
	expensive if either the state-space dimension $\fomdim$ is large, or if the
	time-interval length $T$ is large relative to the time step required to
	numerically integrate Eq.~\eqref{eq:FOM}. For time-critical or many-query
	applications, it is essential to employ an approximation in lieu of the FOM
	ODE~\eqref{eq:FOM}. Projection-based ROMs constitute one such promising
	approach.

\subsection{Reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. For dynamical 
	systems, two distinct types of space--time trial subspaces are commonly used: 
\begin{enumerate} 
	\item \textit{Space–time trial subspaces with spatial dimension reduction (\spatialAcronym)}. These 
	trial subspaces reduce only the spatial dimension of the FOM, and are commonly used
		in model-reduction methods based on spatial projection followed by time
		integration (e.g., Galerkin projection, LSPG projection.)
	\item \textit{Space–time trial subspaces with spatial and temporal dimension reduction (\spaceTimeAcronym)}. These 
	trial subspaces reduce both the spatial and temporal dimensions of the FOM,
		and are commonly employed 
		employed in model-reduction methods based on space--time projection (e.g.,
		space--time LSPG projection).
\end{enumerate}
 We now describe these two types of trial subspaces and their application to 
	Galerkin, LSPG, and space--time LSPG projection.

\subsection{\spatialAcronym\ trial subspaces}
\spatialAcronym\ trial subspaces approximate the FOM ODE solution
	$\stateFOM$ with an approximation $\approxstate$ that resides in an
	affine space--time trial subspace of dimension $K\ll\fomdim$, i.e., 
\begin{equation}\label{eq:spatial_subspace}
\begin{split}
& \stspaceS \defeq \trialspace \otimes \timeSpace +
	\stateIntercept\otimes\onesFunction,\\ 
& \trialspace \defeq \text{Range}(\basismat),  \\%\label{eq:spatial_subspace2}  \\
&\basismat \defeq \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}, %\label{eq:spatial_subspace3} 
\end{split}
\end{equation}
where 
%	$\timeSpace$ denotes the set of real-valued functions acting on the time
%	domain
%	$[0,T]$, 
$\onesFunction\in\timeSpace$ is defined as $\onesFunction:[0,T]\rightarrow\{1\}$,
	$\basismat \in \RRStar{\romdim}{\fomdim}$ denotes the spatial trial subspace with
	$\RRStar{\romdim}{\fomdim}$ denoting the compact Stiefel manifold (i.e.,  $
	\RRStar{\romdim}{\fomdim}\defeq
	\{ \mathbf{X} \in \RR{\fomdim
	\times \romdim}\, \big|\, \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$),
	$\trialspace\subseteq\RR{\fomdim}$ denotes the spatial trial subspace, and
	$\stateIntercept \in \mathbb{R}^{\fomdim}$ denotes the reference state that
	is often taken to be the initial condition (i.e., $\stateIntercept = \stateFOMIC$). 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$, $i=1,\ldots,K$ are typically constructed
using state snapshots, e.g., via
proper orthogonal decomposition (POD)~\cite{berkooz_turbulence_pod} or the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}. 
At any time instance $t\in[0,T]$, ROMs that employ the  \spatialAcronym\ trial subspace approximate the state as
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) = \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$ denotes the generalized
coordinates. Substituting the approximation~\eqref{eq:affine_trialspace} into the FOM ODE~\eqref{eq:FOM} and performing orthogonal
$\elltwo$-projection of the initial condition onto the trial subspace yields
the overdetermined system of ODEs
\begin{equation}\label{eq:g_truncation}
\basisspace \genstateDotArg{}{t} = \velocity(\basisspace
\genstateArg{}{t} + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC, \qquad t \in [0,T].
\end{equation}
Because Eq.~\eqref{eq:g_truncation} is overdetermined, a solution may not
exist. Typically, either \textit{Galerkin projection} or \textit{least-squares
Petrov--Galerkin projection} is employed to reduce the number of equations
such that a unique solution exists. We now describe these two methods.

@@

	\|_{\stweightingMatArg{n}}^2,

@@

time-instantaneous FOM ODE residual evaluated at the state $\basisspace
\genstateGalerkin(t) + \stateIntercept$ and time $t$ over the spatial trial

@@

\subsubsection{Least-Squares Petrov--Galerkin projection}

@@

Space--time methods leverage the \spaceTimeAcronym\ trial subspace $\stspaceST$,
\begin{equation}\label{eq:sttrialspace_def}
 \approxstateArgnt{} \in  \stspaceST \subseteq \RR{N} \otimes \timeSpace,
\end{equation}
and the space--time trial basis \textit{functions},
\begin{align*}
\stbasisArgnt{} &: t \mapsto \stbasis(t) \\
 &: [0,T] \rightarrow  \RR{N \times \stdim} ,%\spaceTrialSpace , \\
%&  \in \RR{N \times \stdim} \otimes \timeTrialSpaceArg{},  
\end{align*}
with $\text{Range}(\stbasis) + \stateInterceptArg{} \equiv \stspaceST$ and where $\stdim$ is the dimension of the space--time generalized coordinates.
%with $\mathbb{V}_{\stdim}( \RR{N} )$ denoting the set of orthogonal $\stdim$ frames in $\RR{N}$ (e.g., an $N \times \stdim$ matrix with orthogonal column vectors).
The state is approximated by,
\begin{equation}\label{eq:stapprox1}
 \stateFOMArg{}{t} \approx \approxstateArg{}{t}  = \stbasisArg{}{t} \stgenstate + \stateInterceptArg{},
\end{equation}
%where $\stbasisArgnt{} : \RR{\stdim} \rightarrow \spaceTrialSpace \otimes \timeTrialSpaceArg{}$ with $ \stbasisArgnt{n} \in \RR{N \times \stdim} \otimes \timeTrialSpaceArg{n}$ is the space--time trial basis, 
where $\stgenstateArg{} \in \RR{\stdim}$ are the space--time generalized coordinates. Critically, the 
space--time basis matrix comprises time-dependent \textit{functions}, while the generalized coordinates are a stationary vector. This is fundamentally different than  
\spatialAcronym\ trial subspaces, in where the generalized coordinates are time-dependent functions while the basis matrix is stationary.

>>>>

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG projection. Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
framework. Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods. Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs.
Section~\ref{sec:numerical_experiments} presents numerical experiments.
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with italicized bold symbols (e.g., $\boldsymbol
x$), vectors with standard bold symbols (e.g., $\mathbf{x}$), 
matrices with capital bold symbols (e.g., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic symbols (e.g., $\mathcal{X}$). We additionally denote differentiation of a time-dependent 
function with respect to with the $\dot\ $ operator.
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical formulation}\label{sec:math}
	We begin by providing the formulation for the full-order model,
	followed by a description of standard model-reduction methods
	classified according to the type of trial subspace they employ.
	\subsection{Full-order model}\label{sec;FOM}
We consider the full-order model to be a dynamical system expressed as a
	system of ordinary differential equations (ODEs)
\begin{equation}\label{eq:FOM}
 \stateFOMDotArg{t}  = \velocity(\stateFOMArg{}{t},t), \qquad \stateFOM(0) =
	\stateFOMIC,\qquad t \in [0,T]
\end{equation}
where  $\stateFOM: [0,T]
	\rightarrow  \RR{\fomdim}$ with $\stateFOM: \timeDummy \mapsto
	\stateFOM(\timeDummy) $ and $\stateFOMDot \equiv {d \stateFOM}/{d\tau}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, 
$T \in \mathbb{R}^+$ denotes the final time, 
 $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial condition,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,\timeDummy) \mapsto
	\velocity(\stateyDiscrete,\timeDummy)$ denotes the velocity, which is possibly
	nonlinear in its first argument. For subsequent exposition, we introduce
	$\timeSpace$ to denote the set of real-valued functions acting on the time
	domain (i.e., $\timeSpace = \{f\,|\,f:[0,T]\rightarrow\RR{}\}$); the
	state can be expressed equivalently as $\stateFOM \in \RR{\fomdim} \otimes
	\timeSpace $.  We refer to the initial value problem defined in
	Eq.~\eqref{eq:FOM} as the ``full-order model" (FOM) ODE. We note that
	although the problem of interest described in the introduction corresponds
	to a parameterized dynamical system, we suppress dependence of the FOM ODE
	\eqref{eq:FOM} on such parameters for notational convenience, as this work
	focuses on devising a projection applicable to a specific parameter instance.

Directly solving the FOM ODE~\eqref{eq:FOM} is computationally expensive if
	either the state-space dimension $\fomdim$ is large, or if the time-interval
	length $T$ is large relative to the time step required to numerically
	integrate Eq.~\eqref{eq:FOM}. For time-critical or many-query applications,
	it is essential to relplace the FOM ODE~\eqref{eq:FOM} with a strategy that
	enables an approximate trajectory to be computed at lower computational
	cost. Projection-based ROMs constitute one such promising approach.

\subsection{Reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. Two types of space--time trial subspaces are commonly used for
	this purpose: 
\begin{enumerate} 
	\item \textit{Subspaces that reduce only the spatial dimension of the full-order
		model (\spatialAcronym)}. These 
	trial subspaces are commonly used
		in model-reduction methods based on spatial projection followed by time
		integration (e.g., Galerkin projection, LSPG projection).
	\item \textit{Subspaces that reduce both the spatial and temporal dimensions of the full-order
		model (\spaceTimeAcronym)}. These trial subspaces are commonly employed in
		model-reduction methods based on space--time projection (e.g., space--time
		LSPG projection).
\end{enumerate}
 We now describe these two types of space--time trial subspaces and their
	application to Galerkin, LSPG, and space--time LSPG projection.

\subsection{\spatialAcronym\ trial subspaces}
\spatialAcronym\ trial subspaces approximate the FOM ODE solution
	$\stateFOM(t)$, $t\in[0,T]$ with an approximation $\approxstate(t)$, $t\in[0,T]$ that resides in an
	affine space--time trial subspace of dimension $K\ll\fomdim$, i.e., 
	$\approxstate\in \stspaceS$, where
\begin{equation}\label{eq:spatial_subspace}
\begin{split}
& \stspaceS \defeq \trialspace \otimes \timeSpace +
	\stateIntercept\otimes\onesFunction.\\ 
\end{split}
\end{equation}
Here, $\trialspace\subseteq\RR{\fomdim}$ with $\dim(\trialspace) = K$ denotes
the spatial trial subspace spanned by an orthogonal basis such that
$ \trialspace= \Range{\basismat}$
with 
$ \basismat \equiv \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}
	\in \RRStar{\romdim}{\fomdim}$, where $\RRStar{\romdim}{\fomdim}$ denotes the compact Stiefel manifold (i.e.,  $
	\RRStar{\romdim}{\fomdim}\defeq
	\{ \mathbf{X} \in \RR{\fomdim
	\times \romdim}\, \big|\, \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$).
%	$\timeSpace$ denotes the set of real-valued functions acting on the time
%	domain
%	$[0,T]$, 
We have also introduced $\onesFunction\in\timeSpace$ as the function $\onesFunction:\timeDummy\mapsto 1$
and
	$\stateIntercept \in \mathbb{R}^{\fomdim}$ as the reference state, which
	is often taken to be the initial condition (i.e., $\stateIntercept = \stateFOMIC$). 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$, $i=1,\ldots,K$ are typically constructed
using state snapshots, e.g., via
proper orthogonal decomposition (POD)~\cite{berkooz_turbulence_pod}, the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}. 
At any time instance $t\in[0,T]$, ROMs that employ the  \spatialAcronym\ trial subspace approximate the state as
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) = \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$ with
$\genstate:\timeDummy\mapsto\genstate(\timeDummy)$
denotes the generalized
coordinates. Substituting the approximation~\eqref{eq:affine_trialspace} into the FOM ODE~\eqref{eq:FOM} and performing orthogonal
$\elltwo$-projection of the initial condition onto the trial subspace yields
the overdetermined system of ODEs
\begin{equation}\label{eq:g_truncation}
\basisspace \genstateDotArg{}{t} = \velocity(\basisspace
\genstateArg{}{t} + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC,
	\qquad t \in [0,T],
\end{equation}
where $\genstateDot\equiv {d \genstate}/{d\tau}$.
Because Eq.~\eqref{eq:g_truncation} is overdetermined, a solution may not
exist. Typically, either \textit{Galerkin projection} or \textit{least-squares
Petrov--Galerkin projection} is employed to reduce the number of equations
such that a unique solution exists. We now describe these two methods.

@@

	\|_{\stweightingMatArg{n}}^2, \qquad \genstate(0) = \genstateIC,
	\qquad t \in [0,T],

@@

FOM ODE residual evaluated at the state $\basisspace
\genstateGalerkin(t) + \stateIntercept$ and time instance $t$ over the spatial trial

@@

\subsubsection{Least-squares Petrov--Galerkin projection}
\KTC{HERE}

@@

Space--time methods leverage the \spaceTimeAcronym\ trial subspace $\stspaceST$,
\begin{equation}\label{eq:sttrialspace_def}
 \approxstateArgnt{} \in  \stspaceST \subseteq \RR{N} \otimes \timeSpace,
\end{equation}
and the space--time trial basis \textit{functions},
\begin{align*}
\stbasisArgnt{} &: t \mapsto \stbasis(t) \\
 &: [0,T] \rightarrow  \RR{N \times \stdim} ,%\spaceTrialSpace , \\
%&  \in \RR{N \times \stdim} \otimes \timeTrialSpaceArg{},  
\end{align*}
with $\Range{\stbasis} + \stateInterceptArg{} \equiv \stspaceST$ and where $\stdim$ is the dimension of the space--time generalized coordinates.
%with $\mathbb{V}_{\stdim}( \RR{N} )$ denoting the set of orthogonal $\stdim$ frames in $\RR{N}$ (e.g., an $N \times \stdim$ matrix with orthogonal column vectors).
The state is approximated by,
\begin{equation}\label{eq:stapprox1}
 \stateFOMArg{}{t} \approx \approxstateArg{}{t}  = \stbasisArg{}{t} \stgenstate + \stateInterceptArg{},
\end{equation}
%where $\stbasisArgnt{} : \RR{\stdim} \rightarrow \spaceTrialSpace \otimes \timeTrialSpaceArg{}$ with $ \stbasisArgnt{n} \in \RR{N \times \stdim} \otimes \timeTrialSpaceArg{n}$ is the space--time trial basis, 
where $\stgenstateArg{} \in \RR{\stdim}$ are the space--time generalized coordinates. Critically, the 
space--time basis matrix comprises time-dependent \textit{functions}, while the generalized coordinates are a stationary vector. This is fundamentally different than  
\spatialAcronym\ trial subspaces, in where the generalized coordinates are time-dependent functions while the basis matrix is stationary.
